Analyzing Fairness: Race, Recidivism, and the COMPAS Algorithm

### 2022-04-21 Paper_Draft2/Draft2.tex

Due to the proprietary nature of the COMPAS algorithm, it is unknown how
exactly these recidivism risk scores are calculated; however, a
\href{https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE\#document/p5/a296598}{sample
COMPAS Risk Assessment Survey} has been made publicly available,
revealing the algorithm's input information. This survey has been widely
critiqued for using proxy variables for race that do not explicitly
factor in a defendant's race but heavily imply it, allowing Northpointe
to claim that their algorithm is free of racial bias. For example, the
COMPAS risk assessment survey asks screeners to speculate if a defendant
might be affiliated with a gang. It also asks if a defendant has any
friends or family members who have been crime victims
\citep{Angwin2016Sample}. Although these questions do not directly ask
about race, they do not take into account the pervasive nature of
systemic racism that infiltrates every aspect of the lives of
marginalized people, thereby indirectly asking about race.
\citet{angwin2016machine}. published a 2016 piece in the news
publication ProPublica that analyzes the methods and algorithms used by
Northpointe in their COMPAS risk score assessment algorithm and uncovers
the racial biases in defendants' scores \citep{angwin2016machine}. Their
research finds that ``the algorithm {[}is{]} somewhat more accurate than
a coin flip,'' a worrisome level of accuracy given the potential impact
its determinations may have on real people's lives.
\citet{angwin2016machine} specifically investigate the distribution of
COMPAS scores by decile among Black and white defendants. They write:
``The analysis also {[}shows{]} that even when controlling for prior
crimes, future recidivism, age, and gender, Black defendants {[}are{]}
45 percent more likely to be assigned higher risk scores than white
defendants'' \citep{larson2016we}. After examining the fairness metric
statistical parity difference, \citet{angwin2016machine} conclude that
the algorithm is racially biased \citep{larson2016we}.

\citet{equivant_response_2016}, on the behalf of Northpointe, deny the
allegations of racial bias and offer their own analyses based on
different fairness metrics in rebuttal \citep{equivant_response_2016}.
\citet{angwin2016machine} maintain that there are biases in the outcome
values, protected attributes, and covariates during
\citet{equivant_response_2016}'s data processing phase. ProPublica
collaborators \citet{larson2016we} account for these biases in their
analyses. In their response, \citet{equivant_response_2016} highlight
that \citet{angwin2016machine} did not account for base rates of
recidivism in their analysis, which are important to understand initial
percentages without the presence of other information.

@@

Our project builds on Women at the Table's various de-biasing algorithms
used in its
\href{https://colab.research.google.com/drive/11SWijBDkr1pSgIXyjUkgoSB0gh5oAlaI}{AI
\& Equality Human Rights Toolbox} to conduct our own analyses on the
COMPAS data set. Based on this analysis, we employ a human rights
framework to contribute to the ProPublica and Northpointe debate and
investigate whether or to what extent there is racial bias in the COMPAS
algorithm. With a solid understanding of the two sides, we aim to
pinpoint the shortcomings of both arguments and correct them in our
analyses. We will use various de-biasing techniques and fairness metrics
to evaluate the level of bias present in the COMPAS data and our
algorithm. We will summarize our results using the JupyterNotebook
framework from Women at the Table, to be used by members of the
organization to teach in a workshop setting. We hope that our findings
will highlight the importance of checking statistical analyses using
varied methods and contribute to the ongoing discussion of the effects
of machine biases in the justice system.

@@

The data we are using for this project is the COMPAS General Recidivism
Risk Scores data set from the AI Fairness 360 (AIF360) toolkit
\citep{aif360-oct-2018}, which does the same initial pre-processing as
ProPublica. The raw data has 6,167 rows and each row represents an
arrest charge for a defendant. AIF360's COMPAS data includes the
defendant's age, race, sex, prior charges, what they were charged with,
and whether or not the defendant ultimately recidivated within a
two-year period after their arrest. For the purposes of our project,
which endeavors to evaluate the differing effects of the COMPAS
algorithm on white defendants versus Black defendants, we have filtered
the data to only include individuals whose race is listed as Caucasian
or African-American. Our data therefore has 5,723 rows (Figure
\ref{fig:table snip}), with the below distributions of race (Figure
\ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year
recidivism rate (Figures \ref{fig:recid plot} \&
\ref{fig:recid race plot} and Table \ref{tab:recid table}).

@@

\caption{This table is a snippet of the data set we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two year recidivism outcome.}\label{fig:table snip}
\end{figure}

@@

\caption{This plot shows the distribution of defendant ages by race. The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants. The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points. The purple dotted line represents the median age of Black defendants (29 years) and the green dotted line represents the median age of white defendants (35 years). For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.}\label{fig:age plot}
\end{figure}

@@

\caption{This plot shows the distribution of two year recidivism outcomes by race. When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate. We can also see that there are more Black defendants in the dataset overall.}\label{fig:recid race plot}
\end{figure}

\begin{longtable}[]{@{}lll@{}}
\caption{This table shows the counts of recidivism by race, illustrating
how a much greater proportion (\textgreater{} 50\%) of Black defendants
recidivated than their white counterparts.
\label{tab:recid table}}\tabularnewline
\toprule
Two Year Recidivism by Race & Recidivated & Survived \\
\midrule
\endfirsthead
\toprule
Two Year Recidivism by Race & Recidivated & Survived \\
\midrule
\endhead
African American & 1661 & 1512 \\
Caucasian & 822 & 1278 \\
\bottomrule
\end{longtable}

@@

@@

\hypertarget{statistical-parity-difference}{%
\subsection{Statistical Parity
Difference}\label{statistical-parity-difference}}

This metric measures the difference that the majority and protected
classes get a particular outcome. The ideal value of this metric is 0.
Fairness for this metric is between -0.1 and 0.1. A negative value means
there is higher benefit for the privileged group (in this case, white
defendants).

\hypertarget{disparate-impact-ratio}{%
\subsection{Disparate Impact Ratio}\label{disparate-impact-ratio}}

This metric is the ratio of how often the favorable outcome occurs in
one group versus the other. In the case of recidivism, this is the ratio
of how many white defendants are predicted to not recidivate compared to
how many Black defendants are predicted to not recidivate. A value of 1
means that the ratio is exactly 1:1. Less than 1 means the privileged
group (white defendants) benefits, while a value greater than 1 means
the unprivileged group (Black defendants) benefits. According to AIF360,
a ratio between .8 to 1.25 is considered fair\citep{Ronaghan2019AI}.

\hypertarget{average-odds-difference}{%
\subsection{Average Odds Difference}\label{average-odds-difference}}

This metric returns the average difference in false positive rate and
true positive rate for the privileged and unprivileged groups. A value
of 0 indicates equality of odds, and a value below 0 implies benefit for
the privileged group. Equality of odds is achieved in the case of
recidivism when the proportion of people who were predicted to
recidivate and did recidivate is equal (true positive rate) for both
Black and white defendants AND the proportion of people who were
predicted to recidivate and did not recidivate (false positive rate) is
equal for both Black and white defendants\citep{aif360-oct-2018}.

\hypertarget{equal-opportunity-difference}{%
\subsection{Equal Opportunity
Difference}\label{equal-opportunity-difference}}

This metric is computed as the difference of true positive rates between
the unprivileged and the privileged groups. The true positive rate is
the ratio of true positives to the total number of actual positives for
a given group\citep{GoogleDev}.

>>>>

Due to the proprietary nature of the COMPAS algorithm, it is unknown how
exactly these recidivism risk scores are calculated. However, a
\href{https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE\#document/p5/a296598}{sample
COMPAS Risk Assessment Survey} has been made publicly available,
revealing the algorithm's input information. \citet{angwin2016machine}
has critiqued for using proxy variables for race that do not explicitly
factor in a defendant's race but heavily imply it, allowing Northpointe
to claim that their algorithm is free of racial bias. For example, the
COMPAS risk assessment survey asks screeners to speculate if a defendant
might be affiliated with a gang. It also asks if a defendant has any
friends or family members who have been crime victims
\citep{Angwin2016Sample}. Although these questions do not directly ask
about race, they do not take into account the pervasive nature of
systemic racism that infiltrates every aspect of the lives of
marginalized people, thereby indirectly asking about race.
\citet{angwin2016machine}. analyzes the methods and algorithms used by
Northpointe in their COMPAS risk score assessment algorithm and uncovers
racial biases in defendants' scores \citep{angwin2016machine}. They find
that ``the algorithm {[}is{]} somewhat more accurate than a coin flip,''
a worrisome level of accuracy given the potential impact its
determinations may have on real people's lives.
\citet{angwin2016machine} specifically investigate the distribution of
COMPAS scores by decile among Black and white defendants. They write:
``The analysis also {[}shows{]} that even when controlling for prior
crimes, future recidivism, age, and gender, Black defendants {[}are{]}
45 percent more likely to be assigned higher risk scores than white
defendants'' \citep{larson2016we}. After examining the fairness metric
statistical parity difference, \citet{angwin2016machine} conclude that
the algorithm is racially biased \citep{larson2016we}.

\citet{equivant_response_2016}, on behalf of Northpointe, deny the
allegations of racial bias and offer their own analyses based on
different fairness metrics in rebuttal \citep{equivant_response_2016}.
\citet{angwin2016machine} maintain that there are biases in the outcome
values, protected attributes, and covariates during
\citet{equivant_response_2016}'s data processing phase. ProPublica
collaborators \citet{larson2016we} account for these biases in their
analyses. In their response, \citet{equivant_response_2016} highlight
that \citet{angwin2016machine} did not account for base rates of
recidivism in their analysis, which are important initial percentages
without the presence of other information.

@@

Our project builds on Women at the Table's various de-biasing algorithms
used in its AI \& Equality Human Rights Toolbox to conduct our own
analyses on the COMPAS data set. Based on this analysis, we employ a
human rights framework to contribute to the ProPublica and Northpointe
debate and investigate whether or to what extent there is racial bias in
the COMPAS algorithm. With a solid understanding of the two sides, we
aim to pinpoint the shortcomings of both arguments and correct them in
our analyses. We will use various de-biasing techniques and fairness
metrics to evaluate the level of bias present in the COMPAS data and our
algorithm. We will summarize our results using the JupyterNotebook
framework from Women at the Table, to be used by members of the
organization to teach in a workshop setting. We hope that our findings
will highlight the importance of checking statistical analyses using
varied methods and contribute to the ongoing discussion of the effects
of machine biases in the justice system.

@@

The data we are using for this project is the COMPAS General Recidivism
Risk Scores data set from the AI Fairness 360 (AIF360) toolkit
\citep{aif360-oct-2018}, which does the same initial pre-processing as
ProPublica. The raw data has 6,167 rows and each row represents an
arrest charge for a defendant. AIF360's COMPAS data includes the
defendant's age, race, sex, prior charges, what they were charged with,
and whether or not the defendant ultimately recidivated within a
two-year period after their arrest. For the purposes of our project,
which endeavors to evaluate the differing effects of the COMPAS
algorithm on white defendants versus Black defendants, we have filtered
the data to only include individuals whose race is listed as Caucasian
or African-American. Our data therefore has 5,273 rows (Figure
\ref{fig:table snip}), with the distributions of race (Figure
\ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two-year
recidivism rate (Figures \ref{fig:recid plot} \&
\ref{fig:recid race plot} and Table \ref{tab:recid table}) shown below.

@@

\caption{A snippet of the data set we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two-year recidivism outcome.}\label{fig:table snip}
\end{figure}

@@

\caption{Distribution of defendant ages by race. The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants. The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points. The purple dotted line represents the median age of Black defendants (29 years) and the green dotted line represents the median age of white defendants (35 years). For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.}\label{fig:age plot}
\end{figure}

@@

\caption{This plot shows the distribution of two-year recidivism outcomes by race. When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate. We can also see that there are more Black defendants in the dataset overall.}\label{fig:recid race plot}
\end{figure}

\begin{longtable}[]{@{}lll@{}}
\caption{Incidence of recidivism by race, illustrating how a much
greater proportion (\textgreater{} 50\%) of Black defendants recidivated
than their white counterparts. \label{tab:recid table}}\tabularnewline
\toprule
Two Year Recidivism by Race & Recidivated & Survived \\
\midrule
\endfirsthead
\toprule
Two Year Recidivism by Race & Recidivated & Survived \\
\midrule
\endhead
African American & 1661 & 1512 \\
Caucasian & 822 & 1278 \\
\bottomrule
\end{longtable}

@@

\textbf{Statistical Parity Difference} - This metric measures the
difference that the majority and protected classes get a particular
outcome. The ideal value of this metric is 0. Fairness for this metric
is between -0.1 and 0.1. A negative value means there is higher benefit
for the privileged group (in this case, white defendants).

\textbf{Disparate Impact Ratio} - This metric is the ratio of how often
the favorable outcome occurs in one group versus the other. In the case
of recidivism, this is the ratio of how many white defendants are
predicted to not recidivate compared to how many Black defendants are
predicted to not recidivate. A value of 1 means that the ratio is
exactly 1:1. Less than 1 means the privileged group (white defendants)
benefits, while a value greater than 1 means the unprivileged group
(Black defendants) benefits. According to AIF360, a ratio between .8 to
1.25 is considered fair\citep{Ronaghan2019AI}.

\textbf{Average Odds Difference} - This metric returns the average
difference in false positive rate and true positive rate for the
privileged and unprivileged groups. A value of 0 indicates equality of
odds, and a value below 0 implies benefit for the privileged group.
Equality of odds is achieved in the case of recidivism when the
proportion of people who were predicted to recidivate and did recidivate
is equal (true positive rate) for both Black and white defendants AND
the proportion of people who were predicted to recidivate and did not
recidivate (false positive rate) is equal for both Black and white
defendants\citep{aif360-oct-2018}.

\textbf{Equal Opportunity Difference} - This metric is computed as the
difference of true positive rates between the unprivileged and the
privileged groups. The true positive rate is the ratio of true positives
to the total number of actual positives for a given
group\citep{GoogleDev}.

@@

### 2022-04-21 final_metrics_analysis.Rmd

```{python crosstable of recitivism by race}
crosstable=pd.crosstab(df_viz['race'],df_viz['two_year_recid'])
crosstable
```

```{python exploratory recidivism mosaic plot}
props={}
props[('Caucasian','Recidivated')]={'facecolor':'red', 'edgecolor':'white'}
props[('Caucasian','Survived')]={'facecolor':'red', 'edgecolor':'white'}
props[('African-American','Recidivated')]={'facecolor':'xkcd:aqua','edgecolor':'white'}
props[('African-American','Survived')]=        {'facecolor':'xkcd:aqua','edgecolor':'white'}
labelizer=lambda k:{('Caucasian','Recidivated'):822,('African-American','Recidivated'):1661,('Caucasian','Survived'):1278,('African-American','Survived'):1512}[k]
mosaic(df_viz,['race','two_year_recid'],labelizer=labelizer,properties=props)
plt.show()
```

@@

@@

```{python exploratory recidivism by race table}
# table of recidivism by race
pd.crosstab(index = df_viz["race"], columns = df_viz["two_year_recid"])
```

```{python exploratory priors by race}
priors_by_race = sns.countplot(x="race", color="priors_count", data=df_viz, palette=workshop_palette)

plt.title('Prior Charges Count by Race')
plt.show()
```

```{python exploratory priors by sex}
priors_by_race = sns.countplot(x="sex", color="priors_count", data=df_viz, palette=workshop_palette)

plt.title('Prior Charges Count by Sex')
plt.show()
```

If a defendant has a higher number of prior offenses, the individual is more likely to be rated with a higher risk of recidivism **do we know this for sure?**. In the former plot, we can see that Black defendants have a higher number of prior charges than do white defendants. This again likely reflects the over-policing of Black people and communities. In the latter plot, we can see that male defendants have a higher number of prior charges than do female defendants. This is worth noting because the COMPAS risk assessment algorithm has been found to express a discrepancy in prediction between sexes **who found this?**. When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes**where does this come from?**.

>>>>

```{python crosstable of recidivism by race}
crosstable=pd.crosstab(df_viz['race'],df_viz['two_year_recid'])
crosstable
```

```{python exploratory recidivism mosaic plot}
props={}
props[('Caucasian','Recidivated')]={'facecolor':'green', 'edgecolor':'white'}
props[('Caucasian','Survived')]={'facecolor':'green', 'edgecolor':'white'}
props[('African-American','Recidivated')]={'facecolor':'xkcd:purple','edgecolor':'white'}
props[('African-American','Survived')]=        {'facecolor':'xkcd:purple','edgecolor':'white'}
labelizer=lambda k:{('Caucasian','Recidivated'):822,('African-American','Recidivated'):1661,('Caucasian','Survived'):1278,('African-American','Survived'):1512}[k]
mosaic(df_viz,['race','two_year_recid'],labelizer=labelizer,properties=props)
plt.show()
```

@@

```{python wordcloud setup}
from os import path
from PIL import Image
from wordcloud import WordCloud, STOPWORDS, ImageColorGenerator
```

```{python wordcloud}
text = " ".join(charge for charge in df_viz.c_charge_desc)
stopwords = set(STOPWORDS)
stopwords.update(["arrest", "case", "charge"])
wordcloud = WordCloud(stopwords=stopwords, background_color="white").generate(text)
```

```{python wordcloud image}
plt.imshow(wordcloud, interpolation='bilinear')
plt.axis("off")
plt.show()
```

@@

```{python exploratory priors by race}
priors_by_race = sns.displot(x="priors_count", row = "race", col = "sex", hue ="race", data=df_viz, palette=workshop_palette)

#plt.title("Prior Charges Count by Race and Sex)
plt.show()
```

In the former plot, we can see that Black defendants have a higher number of prior charges than do white defendants. This again likely reflects the over-policing of Black people and communities. In the latter plot, we can see that male defendants have a higher number of prior charges than do female defendants. This is worth noting as @larson2016we found that there is discrepancy in COMPAS risk assessment algorithm's prediction between sexes. When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes [@larson2016we].
